---
title: "Assignment 3: Physical Properties of Rivers"
author: "Sophia Bryson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of rivers. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/ydeD5axzCnaNzgss9

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-02-08

## Setup

1. Verify your working directory is set to the R project file. Load the tidyverse, dataRetrieval, lubridate, and lfstat packages. Set your ggplot theme (can be theme_classic or something else).
2. Import a data frame called "MysterySiteDischarge" from USGS gage site 03431700. Import discharge data starting on 1964-10-01 and ending on 2021-09-30. Rename columns 4 and 5 as "Discharge" and "Approval.Code". DO NOT LOOK UP WHERE THIS SITE IS LOCATED. 
3. Build a ggplot of discharge over the entire period of record. 

```{r}
# Check working directory
getwd()

# Load libraries
library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(lfstat)

# Set ggplot theme
theme_set(theme_minimal())
library(wesanderson) #load for color palettes

# Import data
site = "03431700"
dischargeCd = "00060" # discharge (ft3/s)
startDt = "1964-10-01"
endDt = "2021-09-30"
MysterySiteDischarge <- readNWISdv(siteNumbers = site, parameterCd = dischargeCd, 
                                   startDate = startDt, endDate = endDt) %>% 
                        rename(Discharge = X_00060_00003, Approval.Code = X_00060_00003_cd)

# Plot discharge over period of record
MysSiteDischargePlot <- ggplot(MysterySiteDischarge, aes(x = Date, y = Discharge)) + 
                        geom_line() +
                        labs(x = "Date", y = "Discharge (cfs)", title = "Mystery Site")
MysSiteDischargePlot
```

## Analyze seasonal patterns in discharge

4. Add a "WaterYear" and "DayOfYear" column to the data frame. Hint: Use a pipe, and you will need both the lubridate and lfstat packages. Set WaterYear to numeric.
5. Create a new data frame called "MysterySiteDischarge.Pattern" that has columns for Day.of.Year, median discharge for a given day of year, 75th percentile discharge for a given day of year, and 25th percentile discharge for a given day of year. Hint: the summarise function includes `quantile`, wherein you must specify `probs` as a value between 0 and 1.
6. Create a plot of median, 75th quantile, and 25th quantile discharges against day of year. Median should be black, other lines should be gray. 
```{r}
# 4 - add cols
MysterySiteDischarge <- MysterySiteDischarge %>% 
                        mutate(Year = year(Date),
                               WaterYear = as.numeric(as.character(water_year(Date, origin = "usgs"))),
                               DayOfYear = yday(Date))

# 5 - new summary df
MysterySiteDischarge.pattern <- MysterySiteDischarge %>%
                                group_by(DayOfYear) %>%
                                summarise(Median.Discharge = median(Discharge, na.rm = TRUE), 
                                          p75.Discharge = quantile(Discharge, 0.75, na.rm = TRUE),
                                          p25.Discharge = quantile(Discharge, 0.25, na.rm = TRUE))

# 6 - plot
DOYDischarge <- ggplot(MysterySiteDischarge.pattern, aes(x = DayOfYear)) +
                       geom_line(aes(y = Median.Discharge)) +
                       geom_line(aes(y = p75.Discharge), color = "gray") +
                       geom_line(aes(y = p25.Discharge), color = "gray") +  
                       labs(x = "Day of Year", y = "Discharge (cfs)") 
DOYDischarge

```

7. What seasonal patterns do you see? What does this tell you about precipitation patterns and climate in the watershed?

> Precipitation is higher from the late fall/early winter through the spring. This seasonal pattern suggests a climatic region with a wet and dry season distinction (rather than snowmelt). 

## Create and analyze recurrence intervals

8. Create two separate data frames for MysterySite.Annual.30yr (first 30 years of record) and MysterySite.Annual.Full (all years of record). Use a pipe to create your new data frame(s) that includes the water year, the peak discharge observed in that year, a ranking of peak discharges, the recurrence interval, and the exceedence probability.

9. Create a plot that displays the discharge vs. recurrence interval relationship for the two separate data frames (one set of points includes the values computed from the first 30 years of the record and the other set of points includes the values computed for all years of the record. 

10. Create a model to predict the discharge for a 100-year flood for both sets of recurrence intervals. 

```{r}
# Data frames with annual peak discharges, recurrence interval, and exceedence probability

recordStart <- min(MysterySiteDischarge$Year) 
startYear <- recordStart #can add one if first year is incomplete and needs to be omitted
year30 <- startYear + 30

MysterySite.Annual.30yr <- MysterySiteDischarge %>%
                           filter(WaterYear >= startYear & WaterYear <= year30) %>% 
                           group_by(WaterYear) %>% 
                           summarise(PeakDischarge = max(Discharge)) %>%
                           mutate(Rank = rank(-PeakDischarge),
                                  RecurrenceInterval = (length(WaterYear) + 1)/ Rank,
                                  ExcProbability = 1/RecurrenceInterval)

MysterySite.Annual.Full <- MysterySiteDischarge %>%
                           group_by(WaterYear) %>% 
                           summarise(PeakDischarge = max(Discharge)) %>%
                           mutate(Rank = rank(-PeakDischarge),
                                  RecurrenceInterval = (length(WaterYear) + 1)/ Rank,
                                  ExcProbability = 1/RecurrenceInterval)

# Plot discharge vs recurrence interval for each df
recurrenceComparison <- ggplot() + 
                        geom_point(data = MysterySite.Annual.30yr, aes(x = RecurrenceInterval, y = PeakDischarge), color = "cadetblue") + 
                        geom_smooth(data = MysterySite.Annual.30yr, aes(x = RecurrenceInterval, y = PeakDischarge), 
                                    method = "lm", color = "cadetblue1", alpha = 0.5) + 
                        geom_point(data = MysterySite.Annual.Full, aes(x = RecurrenceInterval, y = PeakDischarge), color = "tomato") +
                        geom_smooth(data = MysterySite.Annual.Full, aes(x = RecurrenceInterval, y = PeakDischarge), 
                                    method = "lm", color = "tomato1", alpha = 0.5) + 
                        labs(x = "Recurrence Interval (years)", y = "Annual Peak Discharge (cfs)")

# Model  100-year flood discharge for both datasets
   # First 30 years of record
   RIModel_30 <- lm(data = MysterySite.Annual.30yr, PeakDischarge ~ log10(RecurrenceInterval)) #modeled relationship
   RIModel_30$coefficients[1] + RIModel_30$coefficients[2]*log10(100) #calculated discharge for 100 year flood 

   # Full length of record
   RIModel_Full <- lm(data = MysterySite.Annual.Full, PeakDischarge ~ log10(RecurrenceInterval))
   RIModel_Full$coefficients[1] + RIModel_Full$coefficients[2]*log10(100) #calculated discharge for 100 year flood 

```

11. How did the recurrence interval plots and predictions of a 100-year flood differ among the two data frames? What does this tell you about the stationarity of discharge in this river?

> The discharge associated with the 100 year flood event is slightly lower (by 770.5 cfs, a 13% decrease compared with the 30-year window) for the full period of record as compared with just the first 30 years of the record. This relatively low magnitude decrease suggests that the discharge has not changed susbstantially and that stationarity appears to remain a valid assumption for this river. While the change could indicate a slight decrease in discharge, it is also sufficiently minor that it could simply be an artifact of taking a longer timescale which captures more of the natural variability inherent in the system.  


