---
title: "Assignment 4: Physical Properties of Rivers"
author: "Sophia Bryson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of rivers. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/futQwtCsyYsZG9nCA

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-02-15

## Setup and Data Processing

1. Verify your working directory is set to the R project file. Load the tidyverse, cowplot, dataRetrieval, lubridate, lfstat, and EcoHydRology packages. Set your ggplot theme (can be theme_classic or something else).
2. Acquire daily mean discharge data for the Bitterroot River in Montana (USGS gage 12344000) and the Nehalem River in Oregon (USGS gage 14299800). Collect the 10 most recent complete water years.
3. Add columns in the dataset for water year, baseflow, and stormflow. Feel free to use either baseflow separation function we used in class. 
4. Calculate annual total discharge, annual baseflow, and annual proportion baseflow for the two sites.

```{r}
# 1. Workspace setup
   # verify working directory
   getwd()
   
   # load packages
   library(tidyverse)
   library(cowplot)
   library(dataRetrieval)
   library(lubridate)
   library(lfstat)
   library(EcoHydRology)
   
   # set ggplot theme
   theme_set(theme_minimal())
   
# 2. Pull data for gage site
   
   #pull parameters
      discharge_par <- "00060" # discharge (ft3/s)
      start_date <- paste0(year(today() - years(11)),"-10-01") #water year starts Oct. 1, 11 years back gives 10 full years
      end_date <- paste0(year(today() - years(1)), "-09-30") #water year ends Sept. 30, last year to get full year
   
   Bitterroot <- readNWISdv(siteNumbers = "12344000", 
                            parameterCd = discharge_par,
                            startDate = start_date,
                            endDate = end_date)
   
   Nehalem <- readNWISdv(siteNumbers = "14299800", 
                            parameterCd = discharge_par,
                            startDate = start_date,
                            endDate = end_date) 
   
# 3. Add columns: water year, baseflow, and stormflow 
   BitterrootDischarge <- Bitterroot %>% mutate(Discharge = X_00060_00003,
                                                ApprovalCd = X_00060_00003_cd,
                                                Year = year(Date),
                                                WaterYear = as.numeric(as.character(water_year(Date))),
                                                Baseflow = baseflow(Discharge),
                                                StormFlow = Discharge - Baseflow) # Using lfstat version!
   
   NehalemDischarge <- Nehalem %>% mutate(Discharge = X_00060_00003,
                                          ApprovalCd = X_00060_00003_cd,
                                          Year = year(Date),
                                          WaterYear = as.numeric(as.character(water_year(Date))),
                                          Baseflow = baseflow(Discharge),
                                          StormFlow = Discharge - Baseflow) # Using lfstat version!

# 4. Calc site stats: annual total discharge, annual baseflow, and annual proportion baseflow
  
   cfs_to_afy = 723.968 #conversion factor for cubic feet per second to acre feet per year
   
   BitterrootDischarge_Summary <- BitterrootDischarge %>%
                                  group_by(WaterYear) %>% 
                                  summarise(Discharge_AF_yr = sum(Discharge) * cfs_to_afy,
                                            Baseflow_AF_yr = sum(Baseflow) * cfs_to_afy,
                                            Stormflow_AF_yr = sum(StormFlow) * cfs_to_afy,
                                            Proportion_BF = Baseflow_AF_yr/Discharge_AF_yr,
                                            Proportion_SF = Stormflow_AF_yr/Discharge_AF_yr) %>%
                                  mutate_if(is.numeric, round, 2)
   
   NehalemDischarge_Summary <- NehalemDischarge %>%
                               group_by(WaterYear) %>% 
                               summarise(Discharge_AF_yr = sum(Discharge) * cfs_to_afy,
                                         Baseflow_AF_yr = sum(Baseflow) * cfs_to_afy,
                                         Stormflow_AF_yr = sum(StormFlow) * cfs_to_afy,
                                         Proportion_BF = Baseflow_AF_yr/Discharge_AF_yr,
                                         Proportion_SF = Stormflow_AF_yr/Discharge_AF_yr) %>%
                               mutate_if(is.numeric, round, 2)


```

## Analyze seasonal patterns in discharge

5. For both sites, create a graph displaying discharge and baseflow by date. Adjust axis labels accordingly.
6. For both sites, create a graph displaying annual total discharge and annual baseflow across years, and a second graph displaying the proportion baseflow across years (adjust axis labels accordingly). Plot these graphs on top of one another using `plot_grid`. Remember to align the axes!

```{r}
# 5 - graphic display of discharge and baseflow by date
   # combo discharge for comparison
   ggplot() +
      geom_line(data = BitterrootDischarge, aes(x = Date, y = Discharge, color = "Bitterroot River")) +
      geom_line(data = NehalemDischarge, aes(x = Date, y = Discharge, color = "Nehalem River")) + 
      scale_color_manual(name = "", values = c("Bitterroot River" = "darkslategray4", "Nehalem River" = "goldenrod")) +
      theme(legend.position = "bottom") +
      labs(x = "Date", y = "Discharge (cfs)")

   # Bitterroot discharge & baseflow
   ggplot(data = BitterrootDischarge) +
      geom_line(aes(x = Date, y = Discharge, color = "Total Discharge"), size = 1) +
      geom_line(aes(x = Date, y = Baseflow, color = "Baseflow"), size = .8) + 
      scale_color_manual(name = "", values = c("Total Discharge" = "deepskyblue4", "Baseflow" = "indianred3")) +
      theme(legend.position = "bottom") +
      labs(x = "Date", y = "Discharge (cfs)", title = "Bitterroot River")
   
   # Nehalem discharge & baseflow
   ggplot(data = NehalemDischarge) +
      geom_line(aes(x = Date, y = Discharge, color = "Total Discharge"), size = 1) +
      geom_line(aes(x = Date, y = Baseflow, color = "Baseflow"), size = .8) + 
      scale_color_manual(name = "", values = c("Total Discharge" = "deepskyblue4", "Baseflow" = "indianred3")) +
      theme(legend.position = "bottom") +
      labs(x = "Date", y = "Discharge (cfs)", title = "Nehalem River")


# 6 - graphic display of total annual discharge and total annual baseflow, plotted with proportion baseflow across years
   
   #Bitterroot Total Annual Discharge & Baseflow
   Bitterroot_annual_dis <- ggplot(data = BitterrootDischarge_Summary) + 
                              geom_line(aes(x = WaterYear, y = Discharge_AF_yr, color = "Total Annual Discharge")) + 
                              geom_area(aes(x = WaterYear, y = Discharge_AF_yr, fill = "Total Annual Discharge"), alpha = 0.6) + 
                              geom_line(aes(x = WaterYear, y = Baseflow_AF_yr, color = "Total Annual Baseflow")) + 
                              geom_area(aes(x = WaterYear, y = Baseflow_AF_yr, fill = "Total Annual Baseflow"), alpha = 0.6) + 
                              scale_color_manual(name = "", values = c("Total Annual Discharge" = "deepskyblue4", 
                                                                       "Total Annual Baseflow" = "indianred3")) +
                              theme(legend.position = "none") + #cluttering at the moment, so dropping
                              labs(x = "", y = "Discharge (acre feet per year)", title = "Bitterroot River")

   Bitterroot_annual_prop <- ggplot(data = BitterrootDischarge_Summary) + 
                              geom_line(aes(x = WaterYear, y = Proportion_BF), size = 1) + 
                              labs(x = "Water Year", y = "Proportion Baseflow")
   
   Bitterroot_combined <- plot_grid(Bitterroot_annual_dis, Bitterroot_annual_prop, 
                                    ncol = 1, rel_heights = c(1.4, 1), align = "v")
   
   Bitterroot_combined
      
   #Nehalem Total Annual Discharge & Baseflow
   Nehalem_annual_dis <- ggplot(data = NehalemDischarge_Summary) + 
                           geom_line(aes(x = WaterYear, y = Discharge_AF_yr, color = "Total Annual Discharge")) + 
                           geom_area(aes(x = WaterYear, y = Discharge_AF_yr, fill = "Total Annual Discharge"), alpha = 0.6) + 
                           geom_line(aes(x = WaterYear, y = Baseflow_AF_yr, color = "Total Annual Baseflow")) + 
                           geom_area(aes(x = WaterYear, y = Baseflow_AF_yr, fill = "Total Annual Baseflow"), alpha = 0.6) + 
                           scale_color_manual(name = "", values = c("Total Annual Discharge" = "deepskyblue4", 
                                                                    "Total Annual Baseflow" = "indianred3")) +
                           theme(legend.position = "none") + #cluttering at the moment, so dropping
                           labs(x = "r", y = "Discharge (acre feet per year)", title = "Nehalem River")
   
   Nehalem_annual_prop <- ggplot(data = NehalemDischarge_Summary) + 
                           geom_line(aes(x = WaterYear, y = Proportion_BF), size = 1) + 
                           labs(x = "Water Year", y = "Proportion Baseflow")
   
   Nehalem_combined <- plot_grid(Nehalem_annual_dis, Nehalem_annual_prop, 
                                    ncol = 1, rel_heights = c(1.4, 1), align = "v")
   
   Nehalem_combined

   
   
```

7. How do these rivers differ in their discharge and baseflow, both within and across years? How would you attribute these patterns to the climatic conditions in these locations?

> Both rivers display consistent annual cycles of peaks and low flows. In general, the Nehalem River peaks earlier in the year than does the Bitterroot River. The Bitterroot River discharges are, on average, greater in volume than those of the Nehalem.  
The Bitterroot's discharge is, on an annual basis, generally comprised of more baseflow than that of the Nehalem, with baseflow peaking when discharge peaks. This is true to a lesser extent in the Nehalem River, and the Nehalem's proportion of baseflow is more variable across years than that of the Nehalem. 
These trends are congruent with the climatic condition of the two rivers: Located in Montana, the Bitterroot river is subject to inputs from snowmelt, leading to springtime increases in discharge that come in the form of slower release that boost baseflow. The Nehalem, in Oregon, does not have a snowmelt signal, and is instead recieving pulses from increased precipitation/a wet season, meaning that fast-input stormflow contributes a greater proportion relative to the total discharge. 
